---
title: "Reproducible Research:  Peer Assessment 1 - Analyzing Data from FitBit"
author: "Brendan McDougall"
date: "Thursday, April 16, 2015"
output: html_document
keep_md: true
---

## Loading and preprocessing the data
Annonymous personal data from a Fit Bit are analyzed with R.  During a 2-month sampling period, the number of steps taken during a 5-minute interval are recorded.  Experimental data are obtained from: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K].

Date / time values are managed using the R-package *Lubridate* for which [Dates and Times Made Easy with lubridate](http://www.jstatsoft.org/v40/i03/paper) provides good description.  Data tables are manipulated using the R-package *dplyr* for which [Package 'dplyr'](http://cran.r-project.org/web/packages/dplyr/dplyr.pdf) provides description.

The referenced libraries along with other necessary libraries are loaded into R-Studio with commands:
```{r libraryImport, cache=FALSE, results='hide', cache=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(lubridate)
library(lattice)
library(xtable)
```

Subsequently, data are loaded and transformed from experimental data into analytic data using chained (%>%) commands:
```{r dataLoad}
fitBitTblDF <- tbl_df(
    read.csv2("activity.csv", sep = ",", stringsAsFactors = FALSE, na.strings = "NA",
              header = TRUE, # colClasses = c("character", "numeric", "numeric")
    )
)  %>%
    mutate(
        min = formatC(interval, big.mark=":", big.interval=2, width=4, flag = "0"),
        UTCdate = ymd( paste(date)) + hm(min),
        dayNum = yday(UTCdate),
        wDay = wday(UTCdate),
        hourDec = hour(UTCdate) + minute(UTCdate)/60 + second(UTCdate)/3600
    ) %>%
    select(
        UTCdate, min, dayNum, wDay, hourDec, interval, steps
    )%>%
    filter(
        dayNum != 275 & dayNum != 335
        & dayNum != 282 & dayNum != 306 & dayNum != 309 & dayNum != 314 & dayNum != 315 & dayNum != 319
    )
```
For this first order albeit iterated analysis, eight days for which there are no data (all entries = "NA") are excluded using the *filter* command. With *print* command, one reviews data before proceeding with statistical analyses:
```{r dataPrint, echo=FALSE}
print(fitBitTblDF)
```




## What is mean total number of steps taken per day?
Analytic data are processed using chained commands from the R-package *dplyr*:
```{r daySummary}
fitBitSummaryDayNum <-
    fitBitTblDF %>%
    group_by(dayNum) %>%
    summarize(
        count = n(),
        meanIntervalSteps = mean(steps, na.rm = TRUE),
        cumDailySteps = sum(steps, na.rm = TRUE)
    ) %>%
    filter(cumDailySteps != 0) %>%
    arrange(dayNum)
```
A review of processed analytic data shows:
```{r daySumOut, echo=FALSE}
print(fitBitSummaryDayNum)
```
Aside: with the R-package *lubridate*, the UTC date is converted to show the number of day within the calendar year

For each day of the sampling period, processed data lists the mean step-count for the 5-minute intervals by day.  (During a day, there are 288 5-minute intervals, so the processed data shows self-consistency.)  Processed data also show the cumulative number of steps taken each day.  NOTE:  missing data are excluded from the mean of interval step-count and cumulative daily step-count.

From the processed data, the first figure shows a histogram of cumulative daily step-count during the 61-day sampling period from 2012 Oct 1 through 2012 Nov 30.  The *type* for this histogram is set to *count*, so the histogram is of frequency.  Therefore, the bins count the number of days with a particular step-count.  Recall, eight days are excluded from the histogram because no data are recorded.
```{r histogram, echo=FALSE, fig.align='center'}
fitBitSummaryMin <-
    fitBitTblDF %>%
    group_by(hourDec) %>%
    summarize(
        count = n(),
        meanIntervalStepsbyInterval = mean(steps, na.rm = TRUE),
        sdIntervalStepsbyInterval =  sd(steps, na.rm = TRUE)
    ) %>%
    filter(meanIntervalStepsbyInterval != 0) %>%
    arrange(hourDec)

par(
    mfrow = c(1, 1),        # explicitly set plot device to draw 1 graph
    mar = c(4, 4, 2, 1),    # explicitly set margins around plot to default class values
    oma = c(1, 1, 1, 1)     # explicitly set outer margin such that a line of text could be added
)

print(
    histogram(
        ~ cumDailySteps, data = fitBitSummaryDayNum, type = "count",
        main = "Activity:  Tracking Steps vs Time with Fit Bit",
        ylab="Days [#]", xlab = "Steps per Day [#]"
    )
)
```


### Six Number Summary
From the processed data, a Six Number Summary (*summary*) is generated that report min, max, quantiles, mean, and median characteristics.
```{r sixSum, echo=FALSE}
summaryNumbersByDay <- summary(select(fitBitSummaryDayNum,cumDailySteps))
medianDailySteps <- summaryNumbersByDay[3]
meanDailySteps <- summaryNumbersByDay[4]
```
The Six Number summary is displayed using the R-package *xtable*:
```{r echo=F, showtable, results='asis'}
xTbl <- xtable(summaryNumbersByDay)
print(xTbl, type="html")
```
From the Six Number Summary, the mean number of daily steps is **`r meanDailySteps`steps**, and the median number of daily steps is **`r medianDailySteps` steps**.  The reported mean and median are consistent with the histogram.

## What is the average daily activity pattern?
The second figure shows a representative plot of step-counts for a day during the sample period.
```{r plotDay, echo=FALSE,cache=FALSE, fig.align='center'}
par(
    mfrow = c(1, 1),        # explicitly set plot device to draw 1 graph
    mar = c(4, 4, 2, 1),    # explicitly set margins around plot to default class values
    oma = c(1, 1, 1, 1)     # explicitly set outer margin such that a line of text could be added
)

print(
    xyplot(meanIntervalStepsbyInterval ~ hourDec, data = fitBitSummaryMin,
           type = "l", lty=1, col = "black",
           main = "Fit Bit Activity:  Representative Step-Count for a Day",
           ylab="Mean Step Count in 5 Min Interval [#]", xlab = "Hour [#]"
    )
)
```
*Lubridate* is used to convert the UTCdate to decimal hours.  Analytic data are processed such that the plot shows the mean step-count for a given interval averaged across the 53-day data set from within the 61-day sampling period.  The maximum number of step-count is 206 and occurs during the time interval 08:35.



## Imputing missing values
In total, there are 2,304 missing obersations during the sampling period of 61 days.  Recall, eight days for which no data are exluded from the analyses previously reported for a total of 2304 missing observations.  Hence, all the missing data are from this set of 8 days.  Recall, the analysis methodology excluded missing data from within a day for which valid step-count data exists. However, in retrospect, that generality was not needed.

There are technical references that provide guidance for imputing missing values, e.g. [Missing-data imputation](http://www.stat.columbia.edu/~gelman/arm/missing.pdf).  For this exercise, a more pedestrian approach is used to impute missing data while reducing bias.

Missing data are replaced with the mean interval step-count offset by a normal distribution with mean 0 and standard deviation identical to the processed data set.  For example, given that the mean 5-min interval step count is 40 steps with standard deviation 89 steps, a particular instance of missing data could be replaced by the value `r max(0, 40 + rnorm(1, mean = 0, sd = 88))` steps.  (The minimum replacement value is 0 whenever the random sample returns a negative step-count, so dome bias is retained.)  Alternatively, one could randomly select a replacement value from the set of 53 samples for each of the 288 intervals per day.  However, that exercise is left to the reader.


## Are there differences in activity patterns between weekdays and weekends?
